The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-05-26 IEEE SP Magazine/05-2021 - Draft/article.tex
Little update.
Finished Problem section. Added new references.

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is generally associated with the spectral (but also temporal) envelope and is typically assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional timbre feature which is well-adapted to musical data, by deriving it from the constant-Q transform (CQT) \cite{brown1991, brown1992}, a log-scaled frequency transform which matches the notes of the Western music scale. We will show how to decompose the CQT spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these CQT spectral envelope coefficients (CQT-SEC) on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{davis1980}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one single characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2003}, but they mostly resort to using the MFCCs when it comes to having one single timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in a number of MIR tasks, they were originally designed for speech processing applications \cite{davis1980}, using the mel scale, an old perceptual scale built to approximate the auditory system's response \cite{stevens1937}, and are not necessarily adapted to musical data.

%More recently, a number of data-driven approaches are attempting to learn some timbre representation, but can be some embeddings used for music resynthesis and that are not necessarily easily interpretable, we will show that we can use simple signal processing techniques to achieve ... typically, by using autoencoders and generative models
%
% style transfer, music generation,

\cite{engel2017}

@@

>>>>

\IEEEPARstart{T}{imbre} is the attribute of sound which makes, for example, two musical instruments playing the same note sound different. It is generally associated with the spectral (but also temporal) envelope and is typically assumed to be independent from the pitch (but also the loudness) of the sound \cite{moore2004}. In this article, we will show how to design a simple but functional timbre feature which is well-adapted to musical data, by deriving it from the constant-Q transform (CQT) \cite{brown1991, brown1992}, a log-scaled frequency transform which matches the notes of the Western music scale. We will show how to decompose the CQT spectrum into an energy-normalized pitch component and a pitch-independent spectral envelope, the latter from which we will extract a number of timbral coefficients. We will then evaluate the discriminative power of these CQT spectral envelope coefficients (CQT-SEC) on the NSynth dataset \cite{engel2017}, a large-scale dataset of musical notes which is publicly available, comparing them with the mel-frequency cepstral coefficients (MFCCs) \cite{mermelstein1976}, features originally designed for speech recognition but commonly used to characterize timbre in music.

@@

The multidimensional nature of timbre makes it an attribute that is tricky to quantify in terms of one single characteristic feature \cite{grey1977}. While it is assumed to be independent from pitch and loudness, it is not really feasible to fully disentangle timbre from those qualities, as timbre is inherently dependent on the spectral content of the sound which is also defined by its pitch and loudness \cite{moore2004}. Researchers in MIR proposed a number of descriptors to characterize one or more aspects of timbre \cite{peeters2011}, but they mostly resort to using the MFCCs when it comes to having one single timbre feature \cite{mueller2007}. While the MFCCs were shown to be helpful in a number of MIR tasks, they were initially designed for speech processing applications \cite{mermelstein1976} and are not necessarily adapted to musical data. In particular, they are computed through an old process which makes use of the mel scale, a perceptual scale experimentally developed 80 years ago to approximate the human auditory system's response \cite{stevens1937}. More recently, a number of data-driven approaches attempted to learn some timbral representations from musical data, but generally in terms of implicit embeddings which are tied to a specific trained model \cite{engel2017}, \cite{pons2017}, and not necessarily as in explicit and interpretable features such as the MFCCs, which are still usually preferred as the go-to single feature to characterize timbre in music.

@@

%We proposed ...
%
%simple to compute
%interpretable
%functional
%
%We do not claim that it is perfect, but it's a better approximation
